Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models

### 2014-11-18 tex/parametrized-learning.tex
typo, author, acknowledgements

\title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
%\title{Calibrated, Parametrized Learning}
\author{Kyle Cranmer and Daniel Whiteson}
%\date{}							% Activate to display a given date or no date

@@

\begin{itemize}
 \item $x$: a vector of features
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
 \item $\theta$: parameters of a statistical model
\item $f(x| \theta)$:  probability density  (statistical model) for $x$ 
%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
%\item $f( s | \theta )$ The probability density function for $s$ implied by $f(x|\theta)$ and $s(x)$
\item $g( s(x; \theta_0, \theta_1) | \theta )$: The probability density  for $s$ implied by $f(x|\theta_0, \theta_1)$ and $s(x;\theta_0, \theta_1)$
\end{itemize}
We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.

@@

Our main result is that one can form an equivalent test based on 
%\begin{equation}
%T'(D) = \prod_{e=1}^n \frac{ f(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ f(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
%\end{equation}
\begin{equation}\label{eq:equivLRtest}
T'(D) = \prod_{e=1}^n \frac{ g(s_e | \theta_0)}{ g(s_e | \theta_1)}
\end{equation}
if 
\begin{equation}
s_e = s(x_e; \theta_0, \theta_1) = \frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \; 
\end{equation}
or some monotonic function of $s$. This will be proven below.
This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
used to learn $s(x; \theta_1, \theta_0)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$.

While the original goal for hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, the machine learning problem is an event-by-event classification problem. Of course, this follows from the fact that we assume the $x_e$ to be i.i.d.

@@

 The target hypothesis test is be based on 
\begin{equation}
\ln T =   \sum_{e=1}^n \underbrace{\log \left[ \frac {f(x_e | \theta_0) }{ f(x_e | \theta_1) } \right]}_{q(x_e)} \;.
\end{equation}
Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$. A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.

@@

It is not obvious that if we are working with the distributions $g(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators. 
Fortunately, there is a construction based on $g(s|\theta)$ that works. The maximum likelihood estimate is the same as the value that maximizes the ratio to $f(D|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use Theorem~1 to find
\begin{equation}
\hat{\theta} = \argmax_\theta \frac{ f(D | \theta)}{ f(D | \theta_1)} = \argmax_\theta  \sum \ln \frac{f(x_e | \theta)}{f(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{g(s(x_e; \theta, \theta_1) | \theta)}{g(s(x_e; \theta, \theta_1) |\theta_1)} \; .
\end{equation}
It is important that we include the denominator $g(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that change as we vary $\theta$.

@@

\section*{Acknowldgements}
We would like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
project early in its conception and Bal\'azs K\'egl for discussions about the Kaggle challenge and 
non-standard loss functions. KC would like to thank in particular Alex Ihler for challenging discussions that 
led to a reformulation of the initial idea and Shimon Whiteson for patient feedback in that process. 
KC would also like to thank Yuri Shirman for reassuring discussions on elementary vector calculus.
KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
KC would like to thank UC-Irvine for their hospitality while this research was carried out and the 
Moore and Sloan foundations for their generous support of the data science environment at NYU.

>>>>

\title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
%\title{Calibrated, Parametrized Learning}
\author{Kyle Cranmer}
%\date{}							% Activate to display a given date or no date

@@

\begin{itemize}
 \item $x$: a vector of features
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
 \item $\theta$: parameters of a statistical model
\item $f(x| \theta)$:  probability density  (statistical model) for $x$ 
%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
%\item $f( s | \theta )$ The probability density function for $s$ implied by $f(x|\theta)$ and $s(x)$
\item $g( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $f(x|\theta_0, \theta_1)$ \end{itemize}
We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.

@@

Our main result is that one can form an equivalent test based on 
%\begin{equation}
%T'(D) = \prod_{e=1}^n \frac{ f(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ f(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
%\end{equation}
\begin{equation}\label{eq:equivLRtest}
T'(D) = \prod_{e=1}^n \frac{ g(s_e | \theta_0)}{ g(s_e | \theta_1)}
\end{equation}
if 
\begin{equation}
s_e = s(x_e; \theta_0, \theta_1) = \frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \; 
\end{equation}
or some monotonic function of that ratio. This will be proven below.
This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
used to learn $s(x; \theta_1, \theta_0)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$.

While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, the machine learning problem is an event-by-event classification problem. Of course, this follows from the fact that we assume the $x_e$ to be i.i.d.

@@

 The target hypothesis test is based on 
\begin{equation}
\ln T =   \sum_{e=1}^n \underbrace{\log \left[ \frac {f(x_e | \theta_0) }{ f(x_e | \theta_1) } \right]}_{q(x_e)} \;.
\end{equation}
Here we see that the optimal $T$ for the experiment is composed of a sum over events of a  linear function of the per-event function $q(x)$. A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.

@@

It is not obvious that if we are working with the distributions $g(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators. 
Fortunately, there is a construction based on $g(s|\theta)$ that works. The maximum likelihood estimate is the same as the value that maximizes the ratio with respect to $f(D|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use Theorem~1 to find
\begin{equation}
\hat{\theta} = \argmax_\theta \frac{ f(D | \theta)}{ f(D | \theta_1)} = \argmax_\theta  \sum \ln \frac{f(x_e | \theta)}{f(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{g(s(x_e; \theta, \theta_1) | \theta)}{g(s(x_e; \theta, \theta_1) |\theta_1)} \; .
\end{equation}
It is important that we include the denominator $g(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that change as we vary $\theta$.

@@

\section*{Acknowldgements}
KC would like to thank in particular 
Daniel Whiteson for discussions and encouragement throughout the project, 
Alex Ihler for challenging discussions that 
led to a reformulation of the initial idea, and Shimon Whiteson for patient feedback in that process. 
KC would also like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and 
non-standard loss functions, and Yuri Shirman for reassuring discussions on elementary vector calculus.
KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
KC is grateful to UC-Irvine for their hospitality while this research was carried out and the 
Moore and Sloan foundations for their generous support of the data science environment at NYU.
